Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-04-09 article.tex
Improve description part

\richel{Up until here}

The third step infers the actual Bayesian posteriors from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference models from the experiments selected in the previous step. 
For each selected experiment, one BEAST2 posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package.

The fourth step compares the true tree to each posterior tree, resulting in an 
error distribution (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The nLTT statistic (\cite{janzen2015approximate}) is used by default, but also 
a user-defined error statistic can be used. 
As an example, \verb;pirouette; supplies one custom error statistic,
that uses an absolute difference in the gamma statistic 
[\cite{pybus2000testing}].
Additionally, the user can specify the proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of the state space.

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree, following a parallel 
pipeline. We want to produce such tree 'converting' the original tree 
$\tau_{\mathit{P}}$, created according to the generative prior $\mathit{P}$, to 
be in accordance with the chosen inference model $\mathit{I}$:
\begin{align}
    \tau_{\mathit{P}} = (\Vec{t}_{\mathit{P}}, \psi_{\mathit{P}}) 
\xrightarrow[]{\mathit{T}} \tau_{\mathit{I}} = (\Vec{t}_{\mathit{I}}, 
\psi_{\mathit{I}}),
\end{align}
where $\Vec{t}$ are the tree branching times and $\psi$ is tree topology, which 
provides information on where new lineages branch out at each branching time.

To execute the twinning process we exploit the likelihood function 
$L_{\mathit{I}}$ associated with the chosen inference model $\mathit{I}$. We 
find the parameters $\theta^{*}_{\mathit{I}}$ (e.g. speciation and extinction 
rates, in case of a birth-death model) that maximize $L_{\mathit{I}}$ applied 
to the true tree, conditioned on its number of tips $n_{\mathit{P}}$,
\begin{align}
    max[L_{\mathit{I}}(\theta_{\mathit{I}}|\tau_{\mathit{P}}, n_{\mathit{P}})] 
\rightarrow \theta^{*}_{\mathit{I}}.
\end{align}
We use $\theta^{*}_{\mathit{I}}$ to simulate an equal amount $n_{\mathit{I}} = 
n_{\mathit{P}}$ of branching times $\Vec{t}_{\mathit{I}}$ for the twin tree 
$\tau_{\mathit{I}}$, under the process $\mathit{I}$. We then impose that 
$\tau_{\mathit{P}}$ and $\tau_{\mathit{I}}$ must respect the same topology
\begin{align}
    \psi_{\mathit{I}} = \psi_{\mathit{P}}.
\end{align}
Using a twin tree serves as a control: both $\tau_{\mathit{P}}$ and 
$\tau_{\mathit{I}}$ will produce an error distribution. We assume that these 
errors are a natural product of the input tree and the intermediate steps of 
the pipeline. All the factors that can cause such errors are: the branching 
times $\vec{t}$ of the tree (as a consequence of the tree prior), its topology 
$\psi$, the stochasticity induced by simulating the alignment (step 1a 
$\rightarrow$ 2a in Fig.~\ref{fig:pipeline}) and the stochasticity induced by 
the Bayesian inference (step 2a $\rightarrow$ 4a in Fig.~\ref{fig:pipeline}). 
By imposing the same topology and replicating the pipelines a sufficient number 
of times we can sort out all the possible causes that are not a direct 
consequence of the choice of the tree prior.

>>>>

The third step infers the posteriors,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
and those inference models that were selected in the previous step. 
Per selected experiment a posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package. 
This step usually takes up most of the full pipeline's computation time.

The fourth step quantifies the inference error made.
It creates an error distribution, by measuring the difference
between the one true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The default way to quantify the difference between two phylogenies
is the nLTT statistic (\cite{janzen2015approximate}), but any 
user-defined error statistic can be used.
An MCMC needs to 'warm up', as initially it will sample
an unrepresentative part of state space. This phase, called 
the burn-in, is commonly set up to be a small part of the entire run, 
after which it is discarded. By default, \verb;pirouette; 
uses a cut-off value of 10\%.

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
Any phylogeny $\tau$ is the combination of 
branching times $\Vec{t}$ and topology $\psi$.
Starting from a non-standard generative prior $t_p = \mathit{P}$,
we can observe a given tree $\tau_{\mathit{P}}$,
with branching times $\Vec{t}_{\mathit{P}}$ and 
topology $\psi_{\mathit{P}}$, 
but we cannot use the tree prior in Bayesian inference.
The twinning process, T, in accordance with a chosen inference 
model $t_p = \mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving 
topology $\psi_{\mathit{P}}$:

\begin{align}
  \tau_{\mathit{P}} = (\Vec{t}_{\mathit{P}}, \psi_{\mathit{P}}) 
  \xrightarrow[]{\mathit{T}} 
  \tau_{\mathit{I}} = (\Vec{t}_{\mathit{I}}, \psi_{\mathit{P}})
\end{align}

To achieve this, we find the parameters $\theta^{*}_{\mathit{I}}$ 
(e.g. speciation and extinction rates, in case of a birth-death model) 
that maximize likelihood $L_{\mathit{I}}$ applied 
to the true tree, conditioned on its number of tips $n_{\mathit{P}}$,
where $L_{\mathit{I}}$ is  the likelihood function  
associated with the chosen inference model $\mathit{I}$:

\begin{align}
    max[L_{\mathit{I}}(\theta_{\mathit{I}}|\tau_{\mathit{P}}, n_{\mathit{P}})] 
\rightarrow \theta^{*}_{\mathit{I}}.
\end{align}

We use $\theta^{*}_{\mathit{I}}$ to simulate an equal amount 
$n_{\mathit{I}} = n_{\mathit{P}}$ 
of branching times $\Vec{t}_{\mathit{I}}$ for the twin tree 
$\tau_{\mathit{I}}$, under the process $\mathit{I}$, 
while preserving the same topology.

Using a twin tree serves as a control: both $\tau_{\mathit{P}}$ and 
$\tau_{\mathit{I}}$ will produce an error distribution. 
These errors are unavoidably accumulated in the \verb;pirouette; pipeline. 
All the factors that can cause such errors are: the branching 
times $\vec{t}$ of the tree (as a consequence of the tree prior), its topology 
$\psi$, the stochasticity induced by simulating the alignment (step 1a 
$\rightarrow$ 2a in Fig.~\ref{fig:pipeline}) and the stochasticity induced by 
the Bayesian inference (step 2a $\rightarrow$ 4a in Fig.~\ref{fig:pipeline}). 
By imposing the same topology and replicating the pipelines a sufficient number 
of times we can sort out all the possible causes that are not a direct 
consequence of the choice of the tree prior.
